Expectations bias moral evaluations

### 2018-07-18 SuppInfo.Rmd
submitted to PNAS

---
title: "Supporting Information"
output:
  pdf_document: default
  html_notebook: default
---

@@

df <- df_wide %>%
  gather(trial, response, Earthquake:Rape) %>%
  mutate(response = ifelse(trial %in% reverse_items, 6-response, response))
```
# Item Norming

We created and normed 20 items, each consisting of a pair of events that differed in their prior likelihood. The two events (outcomes) were manipulated so what one would see more expected than the other. Participants (_N_ = `r nrow(df_wide)`) were asked to choose which event seemed more likely using a 5-point scale (1 = Event 1 seems a lot more likely, 2 = Event 1 seems a little more likely, 3 = Neither event seems more likely, 4 = Event 2 seems a little more likely, 5 = Event 2 seems much more likely). As predicted, for all items the "more expected" outcome was judged to be more likely. Plot below shows mean likelihood rating by item along with approximate 95% confidence intervals.

>>>>

---
title: "Supporting Information"
output:
  pdf_document:
    keep_tex: true
  # html_notebook: default
---

@@

df <- df_wide %>%
  gather(trial, response, Earthquake:Rape) %>%
  mutate(response = ifelse(trial %in% reverse_items, 6-response, response))
```

# Item Norming

We created and normed 20 items, each consisting of a pair of events that differed in their prior likelihood. The two events (outcomes) were manipulated so what one would seem more expected than the other. Participants (_N_ = `r nrow(df_wide)`) were asked to choose which event seemed more likely using a 5-point scale (1 = Event 1 seems a lot more likely, 2 = Event 1 seems a little more likely, 3 = Neither event seems more likely, 4 = Event 2 seems a little more likely, 5 = Event 2 seems much more likely). As predicted, for all items the "more expected" outcome was judged to be more likely. Plot below shows mean likelihood rating by item along with approximate 95% confidence intervals.

### 2018-07-18 paper/paper.Rmd
submitted to PNAS

abstract: |
 When we learn of tragic events--a violent killing or a terrorist attack--we are mournful and often driven to action. But our responses to these events can differ substantially depending on the context of the event. When tragedy takes us by surprise, we are shocked and outraged, but when it strikes in unstable regions or rough neighborhoods, we are often resigned to feel that “these things happen.” Here we explain the role that expectations seem to play in moral evaluation by modeling evaluations as the comparison of the state of the world before and after an event. According to this model, expectations influence evaluations through information-theoretic principles: less expected events do more to inform us about the state of the world than do more expected events. Consistent with this theory, in two preregistered studies we found that people’s judgments of morally significant events were affected by the prior probability of that event, so that people were more upset about an event that was unexpected (e.g., a robbery at a clothing store) than an event that was more more probable (e.g., a robbery at a convenience store). This bias threatens to impose a pernicious cycle: People living in poverty, geo-politically unstable regions, and other dangerous circumstances have the greatest need for assistance and concern. However, for these very reasons, it is often unsurprising when they are harmed. Consequently, the influence of expectations on moral evaluations threatens to reduce observer's concern for those people who are most in need.

significance: |
  In two behavioral studies, we found that people's expectations biased their reactions to morally-significant events--people are more upset by surprising harms than more expected harms. This bias appears deeply pernicious: because it is often unsurprising when people living in poverty, dangerous neighborhoods, or unstable regions become victims of crime, natural disasters, or terrorism, observers are likely to experience reduced concern for these victims. Yet people living in disadvantaged circumstances are those most in need of assistance and concern from the world at-large. By blunting people's concern for those victims who are most in need, the bias we have uncovered threatens to increase their suffering and further sustain inequality.

@@

```{r make figures, include=FALSE}
fig1 <- plot(marginal_effects(fit_exp1, "trial"))
fig_exp1 <- fig1$trial$data %>%
  ggplot() +
  aes(x = reorder(trial, -estimate__), y = estimate__, ymin = lower__, ymax = upper__) +
  geom_errorbar(width=.25) +
  geom_point(shape=17, size = 2) +
  coord_cartesian(ylim=c(0,1)) +
  geom_hline(yintercept = .5, linetype = "dashed", color = "grey") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    aspect.ratio = .75,
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
    ) +
  labs(
    x = "Trial",
    y = "Probability of judging less\nexpected more upsetting"
  ) 
# library(gridExtra)
# grid.arrange(ggplotGrob(fig_exp1))

@@

plt_cont <- fig2_cont$trial %>%
  ggplot() +
  aes(x = reorder(trial, -estimate__), y = estimate__, ymin = lower__, ymax = upper__) +
  geom_errorbar(width=0) +
  geom_point(shape=17, size = 1) +
  coord_cartesian(ylim=c(2,4)) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "grey") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    aspect.ratio = 10/16,
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
  ) +
  labs(
    x = "Trial",
    y = '"Upsetting" Judgments'
  )

@@

When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action. But our responses to these events differ depending on the context in which these events occur [e.g., @Trope2010]. In particular, our expectations play an important role in our reactions to events. This seems true of both the mundane--when a friend's loved-one passes, we often ask, "was it sudden?"--the dramatic--a murder in a sleepy town prompts a very different reaction than a similar crime in a major city--and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occurs in peaceful rather than unstable regions. For instance, the 2015 Paris attacks were felt around the world. Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut [@Graham2015]. Likewise, stories of the Virginia Tech massacre, which killed 32 people, wildly overshadowed coverage of attacks killing nearly 200 people in Iraq--one of the bloodiest days since the 2003 invasion [@Greenslade2007]. In each of these situations, the surprise elicited by an event appears to affect people’s reactions to the event. In some contexts, harm is tragic and in others, “these things happen.”

Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify this pattern of judgment, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations. Outside of the domain of morality, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events. There is a special thrill to having one’s expectations exceeded and disappointment when events fail to meet expectations: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985]. Indeed, in a controlled study, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Given a gamble with a 10% chance to win \$30 and 90% chance to win \$0, participants felt little disappointment with the \$0 outcome, but were considerably more excited when they won \$30. Conversely, given a gamble with a 90% chance of winning \$60 and 10% chance of winning \$0, participants were disappointed with the \$0 outcome and showed more muted enjoyment of the \$60 outcome. In fact, in gambles similar to these, people were happier with the smaller unexpected gain than with the larger but more expected gain [also see @Shepperd2002].

The influence of expectations on moral evaluations threatens to produce negative social consequences. Tragedies are tragic, whether expected or not, and failure to recognize that could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo. Do expectations directly bias reactions to morally consequential events, as intuition and anecdote seem to suggest? To demonstrate the plausibility of this suggestion, we develop a theoretical account of how expectations influence evaluations, not as a strange quirk or uniquely human bias, but instead as a fundamental component of how people evaluate events.

Most events of consequence consist of changes--a transition from one state to another. For this reason, we construe event evaluation as a comparison between the states of the world before and after the event. Comparison plays a crucial role in many theories of decision-making, and especially in theories aimed at explaining the role of expectations in evaluation  [e.g., @Bell1985; @Gul1991; @Loomes1986; @Mellers1997]. For instance, in Prospect Theory [@Kahneman1979; @Tversky1992], the utility of a prospect is evaluated by comparison to a current reference point. Our proposal is similar: we argue that expectations can set the reference points against which people compare future outcomes, so that assigning value to an event consists in comparing the state of the world following the event to the state of the world just prior to the event. We further assume that people’s knowledge of the world is uncertain, so we represent the states of the world probabilistically. This means that a person’s model for the state of the world is just their expectations about what has happened, what is happening, and what will happen in the future. One direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence. That is, we evaluate events positively or negatively to the extent that they inform our understanding of the world in positive or negative ways.

One striking implication of this account is that it provides an information-theoretic route to understanding how expectations directly influence evaluation. A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability; low probability events carry more information than high probability events [@Shannon1948]. We propose that people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events. For example, if a bombing occurs in Paris--a less expected location--rather than Lebanon--a more expected location--we learn that the world is more dangerous than we had previously believed. This suggestion is novel in the moral domain, but these points are widely accepted elsewhere. For instance, expectations have long been recognized as fundamental to models of human and animal learning [e.g., @Rescorla1972].

@@

From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Theoretical Formalization). Specifically, it is the difference in the expected utility under different values of $S$, weighted by the inverse of the prior probability of $S$.. [^1]

@@

This conclusion does not depend on the nature of the event in question nor on the utility function over events, but instead holds generally wherever we conceive of evaluation as a holistic comparison between uncertain states of the world. Consequently, this account predicts expectations will equally affect reactions to gambles with clearly specified values and probabilities [@Mellers1997] and real-world events with ill-defined values and where reasoners' expectations are defined by their existing beliefs and world knowledge.

Given the potentially fundamental role expectations play in event evaluation, and the potentially serious moral biases that may result, we sought to test empirically whether people’s evaluations of morally harmful events are affected by their expectations about those events. We conducted two preregistered studies examining the role of expectations in moral evaluations. In both studies, participants were presented with a series of trials where they read brief descriptions of two different events and were asked to indicate which of the two events seemed more upsetting. In these trials, the two events were highly similar, but were manipulated so that they differed in their perceived prior probabilities: one event was more expected and one more unexpected.

The design of these studies was chosen with special consideration to two theoretical points. First, the information-theoretic account we have described makes a solely directional prediction--all else being equal, more surprising events will elicit stronger reactions than less surprising events. Asking participants to compare two similar events allowed us to test this directional prediction. Second, it is the informativeness of events, determined by one's prior expectations of the event, that we predict will shape evaluations. Thus, reasoners should not have to be induced to hold expectations experimentally [as in experiments by @Mellers1997], but instead their own prior beliefs will be sufficient to shape their expectations and affect their evaluations. For this reason, we manipulated the prior probability of events by describing two otherwise identical events occurring in different contexts. We explicitly chose contexts that differed in ways that could not be accounted for by construal-level explanations (@Trope2010) or explanations based on biases towards one’s ingroup (@Brewer1979). This allowed us to test the role of expectations based on world knowledge, as well as providing a more naturalistic set of items.

@@

Consistent with our predictions, manipulating expectations consistently shaped evaluations across a range of morally-significant events. We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same. For instance, people judged that a robbery at a clothing store (unexpected) was more upsetting than a robbery at a convenience store (expected). Furthermore, these differences were large, systematic, and robust. Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize our conclusions to items we did not test [@Judd2012].

![Participants’ responses by item in Study 2. Upper: heatmap plot displaying probability of choosing each response on the 1 to 5 scale across items. Lower: marginal expected response across items, with 95% credible interval calculated from the ordinal regression model. Note the scale has been constrained to show relevant region of results. In both plots, response of 3 indicates indifference, a response below 3 indicates a bias to choose the more expected event as more upsetting, and a response above 3 indicates a bias to choose the less expected event as more upsetting. Similar to Study 1, when participants judged that one of the events was more upsetting, they showed a robust bias toward choosing the less expected event as more upsetting. Trials are sorted by response to aid readability. <span data-label="fig:fig1"></span>](fig2.eps)

@@

There's a sense in which it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision. For instance, an act of terrorism in a peaceful nation might be less expected because of our beliefs about the effectiveness of security forces, the stability of national relationships, and positive relationships among people groups in that nation. Therefore, a surprising act of terrorism casts doubt on our deeper understanding of the geopolitical situation, and thereby leads us to suspect similar acts are more likely to occur in the future.

Yet, however well-grounded the underlying principles, there’s a clear moral danger whenever expectations influence our moral concerns [@Koss2000; -@Janoff-bulman1985]. These effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability. In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles. For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large. However, for these very reasons, it is often unsurprising when harm befalls people living in these circumstances. Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms. The present research identifies this potentially pernicious moral bias and grounds it in information theoretic principles. Hopefully, by further understanding these cognitive dynamics, psychologists and policymakers can begin to find ways to reduce or counteract people's tendency to ignore the plight of those in most need of attention.

@@

We analyzed our data by performing Bayesian estimation using the probabilistic programming language Stan [@Carpenter2017]. We tested our preregistered predictions by computing Bayes Factors (i.e. BF01) using the hypothesis function in the R package brms. Bayes Factors express the ratio of the probability of data under the null hypothesis to the probability of the data under an alternative hypothesis. Larger Bayes Factors indicate that the data are more likely under the null hypothesis (i.e., that the intercept is not different from zero) than the alternative hypothesis (i.e., that the intercept is different from zero), and vice versa. As Bayes Factors can be influenced by prior choices (Gelman, Simpson, & Betancourt, 2017), we performed prior robustness checks to confirm that our conclusions would not depend substantially on the specification of the priors.

@@

We predicted that people would judge that events where unexpected harm occurred were more upsetting than events where expected harm occurred. As indicated in Figure 1, we observed this trend. We estimated the magnitude of this effect by fitting a Bayesian logistic random effects model with participants’ responses as the dependent variable (1 = unexpected event is more upsetting; 0 = expected event is more upsetting) and random intercepts for item and subject, allowing us to generalize both to items and subjects we did not test. The intercept in this model represents the log-odds of selecting the unexpected event as being more upsetting. Thus, by examining the population-level intercept, we can test whether participants were biased toward selecting the unexpected event ($\beta$ > 0), the expected event ($\beta$ < 0), or were unbiased ($\beta$ = 0). We found that people were considerably more likely to think that unexpected events were more upsetting than events that were expected, Intercept = `r round(coefs_exp1$estimate, 3)`, 95% CI [`r round(coefs_exp1$lower, 3)`, `r round(coefs_exp1$upper, 3)`], BF01 `r print_bf(bf_exp1)`.
Figure 1 shows participants' responses broken-down by individual items. Participants' bias toward selecting the unexpected event as more upsetting was consistent across the 16 experimental items.

@@

We predicted that participants would evaluate unexpected events as more upsetting than expected events. We estimated the magnitude of this effect by fitting a cumulative (ordinal) random effects model with participants’ scale responses as the dependent variable (1 to 5) and random intercepts for item and subject. This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or lower. For instance, the second coefficient represents the log-odds participants chose a 2 ("outcome 2 seems slightly more upsetting") or lower on the scale. Similarly, the third intercept coefficient represents the log-odds participants chose a 3 or lower on the scale. To examine the predicted effect, we compared the third intercept coefficient (representing the log-odds of choosing 4 or higher) to the inverse of the second intercept coefficient (thereby representing the log-odds *not* choosing a 3 or higher--i.e., choosing a 1 or 2), allowing us to test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option. This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, BF01 `r print_bf(bf_exp2)` (see supporting information for full model results). We observed this difference consistently across nearly every item (Figure 2).

@@

Evaluation of utility is often examined in the context of decision-making. However, people often generate post hoc evaluations of events for their own sake. For instance, people evaluate events to judge the wisdom of their decisions or the morality of their actions.

@@

Thus, it can be seen under this formalization that the role of expectations in the evaluation of events is fundamental in the sense that it does not depend on any specifics concerning the utility function over states of the world. Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.

>>>>

abstract: |
 When we learn of tragic events--a violent killing or a terrorist attack--we are mournful and often driven to action. But responses to these events can differ substantially depending on the context in which they occur. When tragedy takes us by surprise, we are shocked and outraged, but when it strikes in unstable regions or rough neighborhoods, we are often resigned to feel that “these things happen.” Here we explain the role that expectations seem to play in moral evaluation by modeling evaluations as the comparison of the state of the world before and after an event. According to this theory, expectations influence evaluations through information-theoretic principles: less expected events do more to inform us about the state of the world than do more expected events. Consistent with this theory, in two preregistered studies we found that people’s judgments of morally-significant events were affected by the prior probability of that event. People were more upset about an event that was unexpected (e.g., a robbery at a clothing store) than an event that was more more probable (e.g., a robbery at a convenience store). This bias threatens to impose a pernicious cycle: People living in poverty, geo-politically unstable regions, and other dangerous circumstances have the greatest need for assistance and concern. However, for these very reasons, it is often unsurprising when they are harmed. Consequently, the influence of expectations on moral evaluations threatens to reduce observers’ concern for those most in need.

significance: |
  In two behavioral studies, we found that people's expectations biased their reactions to morally-significant events--people were more upset by surprising harms than more expected harms. This bias threatens to impose a pernicious cycle: People living in poverty, unstable regions, and other dangerous circumstances have the greatest need for assistance and concern from the world at-large. Yet for this very reason, it is often unsurprising when they are harmed. By blunting observers' concern for the disadvantaged, this bias could make people less likely to donate money, to volunteer, or to take political action, thereby increasing inequality and compounding suffering. Recognizing and understanding this bias will suggest ways to mitigate its harmful consequences.

@@

```{r make figures, include=FALSE}
fig1 <- plot(marginal_effects(fit_exp1, "trial"))
fig_exp1 <- fig1$trial$data %>%
  ggplot() +
  aes(x = reorder(trial, -estimate__), y = estimate__, ymin = lower__, ymax = upper__) +
  geom_errorbar(width=.25) +
  geom_point(shape=17, size = 2) +
  coord_cartesian(ylim=c(0,1)) +
  geom_hline(yintercept = .5, linetype = "dashed", color = "grey") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    aspect.ratio = .75,
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
    ) +
  labs(
    x = "Trial",
    y = "Prob. less expected\njudged more upsetting"
  ) 
# library(gridExtra)
# grid.arrange(ggplotGrob(fig_exp1))

@@

plt_cont <- fig2_cont$trial %>%
  ggplot() +
  aes(x = reorder(trial, -estimate__), y = estimate__, ymin = lower__, ymax = upper__) +
  geom_errorbar(width=0) +
  geom_point(shape=17, size = 1) +
  coord_cartesian(ylim=c(2,4)) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "grey") +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    aspect.ratio = 10/16,
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
  ) +
  labs(
    x = "Trial",
    y = 'Upset Judgments'
  )

@@

When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action. But our responses to these events differ depending on the context in which they occur [e.g., @Trope2010]. In particular, our expectations play an important role in our reactions to events. This seems true of the ordinary: when a friend's loved-one passes, we often ask, "was it sudden?"; the dramatic: a murder in a sleepy town prompts a very different reaction than a similar crime in a major city; and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occur in peaceful rather than unstable regions. For instance, the 2015 Paris attacks were felt around the world. Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut [@Graham2015]. Likewise, stories of the Virginia Tech massacre, which killed 32 people, wildly overshadowed coverage of attacks killing nearly 200 people in Iraq--one of the bloodiest days since the 2003 invasion [@Greenslade2007]. In each of these situations, the surprise elicited by an event appears to have affected people’s reactions to it. In some contexts, harm is tragic. In others, “these things happen.”

Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify these judgments, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations. Outside of the domain of morality, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events. There is a special thrill to having one’s expectations exceeded and disappointment when events fail to meet expectations: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985]. Indeed, in a controlled study, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Participants were often more excited by an small unexpected gain (e.g., winning \$30 with 10% probability) than they were by a larger but more expected gain (e.g., winning \$60 with 90% probability) [also see @Shepperd2002].

It seems intuitive that surprise would shape people’s reactions to parties, sporting events, and games of chance. However, it seems wholly inappropriate that surprise would shape people’s reactions to events with serious moral implications. Tragedies are tragic, whether expected or not, and failure to recognize this could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo. If expectations truly do influence our reactions to morally-significant world events, as intuition and anecdote seem to suggest, there are potentially serious social consequences. To demonstrate the plausibility of this suggestion, we develop a theoretical account of how expectations influence evaluations, not as a strange quirk or uniquely human bias, but instead as a fundamental component of how people evaluate events.

Events, as we will conceive of them, are changes in the state of the world--transitions from one state to another. For this reason, we construe event evaluation as a comparison between the states of the world before and after the event. Similar comparison processes play crucial roles in many theories of decision-making [e.g., @Bell1985; @Gul1991; @Loomes1986; @Mellers1997]. For instance, in Prospect Theory [@Kahneman1979; @Tversky1992], the utility of a prospect is evaluated by comparison to a current reference point. Our proposal is similar: we argue that expectations can set the reference points against which people compare future outcomes, so that assigning value to an event consists in comparing the state of the world following the event to the state of the world just prior to the event. Further, we assume that people’s knowledge of the world is uncertain, so we represent the states of the world probabilistically. This means that a person’s model for the state of the world is simply their expectations about what has happened, what is happening, and what will happen in the future. A direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence. That is, we evaluate events positively or negatively to the extent that they inform our understanding of the world in positive or negative ways.

One striking implication of this account is that it provides an information-theoretic route to understanding how expectations directly influence evaluation. A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability; low probability events carry more information than high probability events [@Shannon1948]. We propose that people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events. For example, compared to a bombing in Lebanon, if a bombing occurs in Paris, a less expected location, we learn that the world is more dangerous than we had previously believed. Although this suggestion is novel in the moral domain, these points are widely accepted elsewhere. For instance, expectations have long been recognized as fundamental to models of human and animal learning [e.g., @Rescorla1972].

@@

From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Theoretical Formalization). Specifically, it is the difference in the expected utility under different values of $S$, weighted by the inverse of the prior probability of $S$. [^1]

@@

This conclusion does not depend on the nature of the event in question nor on the utility function over events, but instead holds generally wherever we conceive of evaluation as a holistic comparison between uncertain states of the world. Consequently, this account predicts expectations will equally affect reactions to gambles with clearly specified values and probabilities [@Mellers1997], as well as real-world events with ill-defined values, where reasoners' expectations are defined by their existing beliefs and world knowledge.

Given the potentially fundamental role that expectations play in event evaluation, and the potentially serious moral biases that may result, we sought to test empirically whether people’s evaluations of morally harmful events are affected by their expectations about those events. We conducted two preregistered studies examining the role of expectations in moral evaluations. In both studies, participants were presented with a series of trials in which they read brief descriptions of two different events and were asked to indicate which of the two events seemed more upsetting. In these trials, the two events were highly similar, but were manipulated so that they differed in their perceived prior probabilities: one event was more expected and one more unexpected.

The design of these studies was chosen with special consideration to two theoretical points. First, the information-theoretic account we have described makes a solely directional prediction--all else being equal, more surprising events will elicit stronger reactions than less surprising events. Asking participants to compare two similar events allowed us to test this directional prediction. Second, it is the informativeness of events, determined by one's prior expectations of the event, that we predicted would shape evaluations. Thus, reasoners should not have to be induced to hold expectations experimentally [as in experiments by @Mellers1997], but instead their own prior beliefs should be sufficient to shape their expectations and affect their evaluations. For this reason, we manipulated the prior probability of events by describing two otherwise identical events occurring in different contexts. We explicitly chose contexts that differed in ways that could not be accounted for by construal-level explanations [@Trope2010] or explanations based on biases towards one’s ingroup [@Brewer1999]. This both allowed us to test the role of expectations based on world knowledge and provided a more naturalistic set of items.

@@

Consistent with our predictions, manipulating expectations shaped evaluations across a range of morally-significant events. We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same. For instance, people judged that a robbery at a clothing store (unexpected) was more upsetting than a robbery at a convenience store (expected). Furthermore, these differences were large, systematic, and robust. Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize our conclusions to items we did not test [@Judd2012].

![Participants’ responses by item in Study 2. Upper: heatmap plot displaying probability of choosing each response on the 1 to 5 scale across items. Lower: marginal expected response across items, with 95% credible interval calculated from the ordinal regression model. Note the scale has been constrained to show relevant region of results. In both plots, a response of 3 indicates indifference, responses below 3 indicate a bias to choose the more expected event as more upsetting, and responses above 3 indicate a bias to choose the less expected event as more upsetting. Similar to Study 1, participants showed a robust bias toward choosing the less expected event as more upsetting. Trials are sorted by response to aid readability. <span data-label="fig:fig1"></span>](fig2.eps)

@@

In some contexts, it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision [@Rescorla1972]. For instance, an act of terrorism in a peaceful nation might be less expected because of our beliefs about the effectiveness of security forces, the stability of national relationships, and positive relationships among people groups in that nation. Therefore, a surprising act of terrorism casts doubt on our deeper understanding of the geopolitical situation, and thereby leads us to suspect similar acts are more likely to occur in the future.

Yet, however well-grounded the underlying principles, there is a clear moral danger whenever expectations influence our moral concerns. In particular, these effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability. In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles. For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large. However, for precisely these reasons, it is often unsurprising when harm befalls people living in these circumstances. Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms. The present research identifies this potentially pernicious moral bias and grounds it in information theoretic principles. Hopefully, by further understanding these cognitive dynamics, psychologists and policymakers can begin to find ways to reduce or counteract people's tendency to ignore the plight of those in most need of attention.

@@

We analyzed our data by performing Bayesian estimation using the probabilistic programming language Stan [@Carpenter2017]. We tested our preregistered predictions by computing Bayes Factors (i.e. $BF_{01}$) using the hypothesis function in the R package brms. Bayes Factors express the ratio of the probability of data under the null hypothesis to the probability of the data under an alternative hypothesis. Larger Bayes Factors indicate that the data are more likely under the null hypothesis (i.e., that the intercept is not different from zero) than the alternative hypothesis (i.e., that the intercept is different from zero), and vice versa. As Bayes Factors can be influenced by prior choices [@Gelman2017], we performed prior robustness checks to confirm that our conclusions would not depend substantially on the specification of the priors.

@@

We predicted that people would judge that events where unexpected harm occurred were more upsetting than events where expected harm occurred. As indicated in Figure 1, we observed this trend. We estimated the magnitude of this effect by fitting a Bayesian logistic random effects model with participants’ responses as the dependent variable (1 = unexpected event is more upsetting; 0 = expected event is more upsetting) and random intercepts for item and subject, allowing us to generalize both to items and subjects we did not test. The intercept in this model represents the log-odds of selecting the unexpected event as being more upsetting. Thus, by examining the population-level intercept, we can test whether participants were biased toward selecting the unexpected event ($\beta$ > 0), the expected event ($\beta$ < 0), or were unbiased ($\beta$ = 0). We found that people were considerably more likely to think that unexpected events were more upsetting than events that were expected, Intercept = `r round(coefs_exp1$estimate, 3)`, 95% CI [`r round(coefs_exp1$lower, 3)`, `r round(coefs_exp1$upper, 3)`], $BF_{01}$ `r print_bf(bf_exp1)`.
Figure 1 shows participants' responses broken-down by individual items. Participants' bias toward selecting the unexpected event as more upsetting was consistent across the 16 experimental items.

@@

We predicted that participants would evaluate unexpected events as more upsetting than expected events. We estimated the magnitude of this effect by fitting a cumulative (ordinal) random effects model with participants’ scale responses as the dependent variable (1 to 5) and random intercepts for item and subject. This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or lower. For instance, the second coefficient represents the log-odds participants chose a 2 ("outcome 2 seems slightly more upsetting") or lower on the scale. Similarly, the third intercept coefficient represents the log-odds participants chose a 3 or lower on the scale. To examine the predicted effect, we compared the third intercept coefficient (representing the log-odds of choosing 4 or higher) to the inverse of the second intercept coefficient (thereby representing the log-odds *not* choosing a 3 or higher--i.e., choosing a 1 or 2), allowing us to test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option. This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, $BF_{01}$ `r print_bf(bf_exp2)` (see supporting information for full model results). We observed this difference consistently across nearly every item (Figure 2).

@@

@@

Under this formalization, the role of expectations in the evaluation of events does not depend on any further assumptions about the states of the world or the utility function to be applied over them. Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.
